---
title: "Introduction to testhta"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to testhta}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(testhta)
```

`testhta` provides a software-engineering-inspired framework for building and validating Health Technology Assessment (HTA) cost-effectiveness models in R. By combining cohort Markov models with structured unit tests, it ensures model transparency, correctness, and reproducibility.

This vignette covers:
1. Setting up a Markov model.
2. Running the model under baseline parameters.
3. Accessing model outputs using getters.

## 1. The HTA Markov Model Setup

The core function in `testhta` is `ce_markov()`, which runs a cohort Markov model. The model expects:
- `start_pop`: Initial counts of the cohort across health states.
- `p_matrix`: A 3-dimensional array of transition probabilities (states $\times$ states $\times$ treatments).
- `state_c_matrix`: A matrix of costs for state occupancy.
- `trans_c_matrix`: A matrix of costs incurred during state transitions.
- `state_q_matrix`: A matrix of quality-of-life weights (utilities) for each state.

The package includes a built-in baseline dataset called `test_data`. Let's inspect its structure:

```{r test_data}
# Load baseline dataset
data(test_data)
names(test_data)
```

It contains:
- `start_pop`: Vector of length 3 (Asymptomatic disease, Progressive disease, Dead)
- `p_matrix`: $3 \times 3 \times 2$ array for `without_drug` and `with_drug`
- `state_c_matrix`: State cost matrix for each treatment
- `state_q_matrix`: Utility matrix for each treatment
- `discount_rate`: Standard HTA discount rate (typically 3.5% or 0.035)
- `n_cycles`: Time horizon of the model

## 2. Running the Model

We can run the model by passing `test_data` parameters directly to `ce_markov()`, or using the wrapper helper `run_model()`.

Here is how you execute the Markov simulation:

```{r run_model}
# Run with the helper wrapper
results <- run_model(test_data)
```

The output of `run_model()` is a list containing the population distributions, cycle-specific costs and QALYs, and summary metrics.

## 3. Extracting Results using Getters

`testhta` provides clean getter functions to extract the key outcomes from your model results:

- `get_qalys()`: Extracts total Quality-Adjusted Life Years (QALYs).
- `get_costs()`: Extracts total costs.
- `get_le()`: Extracts total Life Expectancy (undiscounted life-years).
- `get_icer()`: Calculates the Incremental Cost-Effectiveness Ratio (ICER) between the two arms.

Let's retrieve these values from our results:

```{r getters}
# Get total QALYs
get_qalys(results)

# Get total Costs
get_costs(results)

# Get life expectancy
get_le(results)

# Calculate the ICER
get_icer(results)
```

## 4. Parameter Modification with Setters

The package provides setter functions that follow a tidy workflow using the base R pipe (`|>`). These allow you to easily modify input parameters for scenario or sensitivity analyses before running the model:

- `set_discount_rate(input, discount_rate)`
- `set_time_horizon(input, n_cycles)`
- `set_costs(input, scenario, values)`
- `set_utilities(input, scenario, values)`

For example, to calculate the ICER with no discounting and a shorter time horizon (20 cycles):

```{r piping}
scenario_results <- test_data |>
  set_discount_rate(0) |>
  set_time_horizon(20) |>
  run_model()

get_icer(scenario_results)
```

This piping framework makes scenario analysis clean, readable, and less prone to copy-paste errors.
