---
title: "Model Validation and Verification Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Validation and Verification Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(testhta)
library(testthat)
```

Health economic models are critical tools for healthcare decision-making, but their complexity makes them vulnerable to implementation errors. `testhta` provides a structured framework for model verification—ensuring that the mathematical and logical implementation of the model matches its conceptual specification.

Rather than trying to read and manually inspect hundreds of lines of code, this package encourages modelers to write automated **unit tests** using the `testthat` framework to check model behavior under specific bounding conditions.

This guide describes:
1. Core validation rules (test cases).
2. The testing helpers provided by `testhta`.
3. How to write validation tests for your own models.

## 1. Core Model Validation Rules

Several logical rules must hold true for any cost-effectiveness model. `testhta` facilitates writing assertions for these:

- **Discounting**: Total QALYs and costs with standard discounting (e.g., 3.5%) must be strictly lower than undiscounted QALYs and costs.
- **Extreme Discounting**: If the discount rate is extremely high (e.g. $\infty$), all future costs and QALYs should go to zero.
- **Utility Bounding**: If state utilities are set to 0, total QALYs must equal 0. If state utilities are set to 1 and the discount rate is 0, total QALYs must exactly equal life expectancy.
- **Life Expectancy Consistency**: Life expectancy should not be affected by changes to cost or utility values, nor should it change with the discount rate.
- **Absorbing States**: If transition probabilities to the death state are 1, life expectancy must be exactly 1 cycle. If transition probabilities to death are 0, life expectancy must equal the time horizon (`n_cycles`).

## 2. Using `testhta` Validation Helpers

To make writing unit tests as readable and succinct as possible, `testhta` exports several test assertion helpers:

### Bounding Value Checks

These functions check model outcomes against a single expected value:

- `check_model_qalys(expected_qalys, data, ...)`
- `check_model_costs(expected_costs, data, ...)`
- `check_model_le(expected_le, data, ...)`

For example, to test that QALYs are exactly zero when the discount rate is set to 1 (100%):

```{r test_qaly_zero}
data(test_data)

check_model_qalys(
  expected_qalys = 0,
  data = test_data,
  discount_rate = 1,
  label = "QALYs are 0 under 100% discount rate"
)
```

Or that life expectancy is exactly 1 cycle when death probability is 1:

```{r test_le_one}
check_model_le(
  expected_le = 1,
  data = test_data,
  p_healthy_death = 1,
  p_sick_death = 1,
  label = "LE is 1 when death probability is 1"
)
```

### Relational Comparisons

Use `compare_model_runs()` to test that changing a parameter shifts model outcomes in the expected direction:

- **Parameters**:
  - `extractor_fn`: Function to get the outcome (e.g., `get_qalys`, `get_costs`, `get_le`).
  - `comparison_fn`: Comparison expectation (e.g., `expect_lt`, `expect_gt`, `expect_equal`).
  - `params_1`: List of parameter overrides for run 1.
  - `params_2`: List of parameter overrides for run 2.
  - `data`: Base dataset.

To verify that discounting reduces total costs:

```{r test_discount_cost}
compare_model_runs(
  extractor_fn = get_costs,
  comparison_fn = expect_lt,
  params_1 = list(discount_rate = 0.035), # Run 1: discounted
  params_2 = list(discount_rate = 0),     # Run 2: undiscounted
  data = test_data,
  label = "Discounted costs are less than undiscounted costs"
)
```

## 3. Integrating Validation into a Model Development Workflow

When developing or modifying a health economic model, place your test files in the `tests/testthat/` directory of your R package.

For example, create `tests/testthat/test-model_validation.R`:

```r
library(testthat)
library(testhta)

data(test_data)

test_that("economic logic holds for cohort markov model", {
  # Discounting logic
  compare_model_runs(get_qalys, expect_lt, list(discount_rate = 0.035), list(discount_rate = 0), test_data)
  
  # Life Expectancy logic
  check_model_le(expected_le = 50, data = test_data, p_healthy_death = 0, p_sick_death = 0, n_cycles = 50)
  
  # Zero costs logic
  check_model_costs(expected_costs = 0, data = test_data, c_healthy = 0, c_sick = 0, c_intervention = 0, c_death = 0)
})
```

You can execute your tests at any point during development by running:

```r
devtools::test()
```

This automates the verification process and acts as a safety net against accidental bugs when updating transitions, costs, or utility functions.
