---
title: "Validating HTA Models with Unit Tests"
subtitle: "A guide to applying software engineering best practices to health economics"
author: "HTA Modelling Team"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: show
    theme: cosmo
    number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# In a real scenario, you would load your package here:
# library(testhta)
library(testthat)

# --- MOCKING PACKAGE FUNCTIONS FOR THIS DOCUMENT TO RUN STANDALONE ---
# Ideally, these are loaded via library(testhta). We include them here
# so this document renders successfully for demonstration purposes.

# 1. The core model
ce_markov <- function(start_pop, p_matrix, state_c_matrix, trans_c_matrix, state_q_matrix, 
                      n_cycles = 46, init_age = 55, s_names = NULL, t_names = NULL, discount_rate = 0.035) {
  
  # Simplified version of the package function for demonstration
  n_states <- length(start_pop)
  n_treat <- dim(p_matrix)[3]
  
  # Initialize total accumulators
  total_costs <- setNames(rep(0, n_treat), t_names)
  total_QALYs <- setNames(rep(0, n_treat), t_names)
  total_LE <- setNames(rep(0, n_treat), t_names)
  
  for (i in 1:n_treat) {
    # Just a mock calculation to satisfy the tests in this guide
    # In reality, this contains the Markov trace logic
    
    # Logic for Discounting Test:
    base_qaly <- 10 
    if(discount_rate > 0) base_qaly <- base_qaly * (1 / (1+discount_rate))
    if(discount_rate > 0.5) base_qaly <- 0 # Simulating high discount
    
    # Logic for Zero Cost Test:
    cost_mult <- 1
    if(sum(state_c_matrix) == 0) cost_mult <- 0
    
    # Logic for LE:
    le_val <- n_cycles # Default alive
    # If prob death is 1 immediately
    if(p_matrix[1, "Dead", i] == 1) le_val <- 1
    
    total_QALYs[i] <- base_qaly * mean(state_q_matrix[i,])
    total_costs[i] <- 1000 * cost_mult * (1 / (1+discount_rate))
    total_LE[i] <- le_val
  }
  
  list(total_costs = total_costs, total_QALYs = total_QALYs, total_LE = total_LE)
}

# 2. Getters
get_qalys <- function(res) res$total_QALYs
get_costs <- function(res) res$total_costs
get_le <- function(res) res$total_LE
```

# Introduction

In Health Technology Assessment (HTA), errors in model coding can lead to incorrect policy decisions. While we often validate models against literature or clinical data (face validity), we often neglect **verification**—ensuring the code actually does what we think it does.

This guide demonstrates how to use the `testthat` package to perform **black-box testing** on a Markov model. Instead of reading every line of code, we test the model's *behavior* by feeding it specific inputs and asserting that the outputs match our logical expectations.

## The Philosophy

We treat the model as a function $f(x)$. We don't need to look inside the function to test it. We simply define:
1.  **Input ($x$):** A specific set of parameters (e.g., discount rate = 0).
2.  **Expected Output ($y$):** The logical result (e.g., Costs should equal undiscounted costs).
3.  **The Test:** Does $f(x) == y$?

# Setup: The Model Wrapper

To make testing easy, we define a helper function `run_model`. This allows us to pass the default dataset but override specific parameters we want to test (like the discount rate).

```{r wrapper}
# Load basic test data (simulated for this guide)
t_names <- c("without_drug", "with_drug")
s_names <- c("Asymptomatic", "Progressive", "Dead")

# Base input parameters
base_params <- list(
  start_pop = c(1000, 0, 0),
  p_matrix = array(0, dim = c(3,3,2), dimnames = list(s_names, s_names, t_names)),
  state_c_matrix = matrix(c(100, 200, 0, 500, 200, 0), nrow = 2, byrow = TRUE),
  trans_c_matrix = matrix(0, nrow = 3, ncol = 3),
  state_q_matrix = matrix(1, nrow = 2, ncol = 3), # Utility = 1
  n_cycles = 15
)

# Wrapper function
run_model <- function(input_data, ...) {
  # Merge default input_data with manual overrides (...)
  updates <- list(...)
  args_list <- modifyList(input_data, updates)
  
  # Call the main model function
  do.call(ce_markov, args_list)
}
```

::: {.callout-tip}
## Why a wrapper?
Using `...` (ellipses) allows us to write clean tests like `run_model(base_data, discount_rate = 0)` without having to manually reconstruct the entire input list every time.
:::

# Writing Tests

We will now write three types of validity checks based on the `testhta` framework.

## 1. Economic Logic: Discounting

A fundamental property of economic models is that future value decreases as the discount rate increases.

**The Test Case:**
* **A:** Run model with discount rate = 0% (Undiscounted).
* **B:** Run model with discount rate = 3.5% (Discounted).
* **Expectation:** Result B < Result A.

```{r test-discounting}
test_that("Discounting logic holds", {
  
  # 1. Run variations
  res_undiscounted <- run_model(base_params, discount_rate = 0.00)
  res_discounted   <- run_model(base_params, discount_rate = 0.035)
  
  # 2. Extract specific outcomes
  qaly_undiscounted <- get_qalys(res_undiscounted)
  qaly_discounted   <- get_qalys(res_discounted)
  
  # 3. Assertions
  expect_lt(qaly_discounted["with_drug"], qaly_undiscounted["with_drug"])
  
  # Bonus check: High discount rate should drive value near zero
  res_high_disc <- run_model(base_params, discount_rate = 1000) 
  expect_lt(get_qalys(res_high_disc)["with_drug"], 1)
})
```

## 2. Boundary Values: Zero Inputs

These are "sanity checks." If we remove all costs from the inputs, the model must output zero cost. If it doesn't, there is likely a hard-coded value hidden somewhere in the calculation engine.

**The Test Case:**
* **Input:** Set `state_c_matrix` and `trans_c_matrix` to 0.
* **Expectation:** Total costs = 0.

```{r test-zeros}
test_that("Zero inputs result in zero outputs", {
  
  # Create zeroed-out matrices
  zero_state_c <- base_params$state_c_matrix * 0
  zero_trans_c <- base_params$trans_c_matrix * 0
  
  # Run model
  res_zeros <- run_model(base_params, 
                         state_c_matrix = zero_state_c,
                         trans_c_matrix = zero_trans_c)
  
  # Assert costs are exactly 0
  actual_costs <- get_costs(res_zeros)
  expect_equal(as.numeric(actual_costs), c(0, 0))
})
```

## 3. Clinical Logic: Life Expectancy

We can validate the transition matrix logic by forcing extreme mortality risks.

**The Test Case:**
* **Input:** Set probability of death to 1.0 (100%) immediately for everyone.
* **Expectation:** Life Expectancy (LE) should be exactly 1 cycle.

```{r test-clinical}
test_that("Immediate mortality results in LE of 1", {
  
  # Modify p_matrix to force death
  dead_p_matrix <- base_params$p_matrix
  dead_p_matrix[, "Dead", ] <- 1 # All transitions go to Dead
  
  res_death <- run_model(base_params, p_matrix = dead_p_matrix)
  
  # Assert LE is 1
  expect_equal(as.numeric(get_le(res_death)), c(1, 1))
})
```

# Automating the Process

While the interactive examples above are useful for debugging, the true power of unit testing comes from automation.

In the `testhta` package structure, these tests live in `tests/testthat/`. You can run all tests at once to ensure that new changes to the model haven't broken existing logic.

### Running all tests

```r
# Run all tests in the package
devtools::test()
```

### Typical Output

When `devtools::test()` is run, you will see a report in the console:

```text
[ FAIL 0 | WARN 0 | SKIP 0 | PASS 3 ]
```

* **PASS:** The model logic is holding up.
* **FAIL:** You broke something! The error message will tell you exactly which logic check failed (e.g., "Discounted QALYs were not less than undiscounted QALYs").

# Summary

By implementing `testhta`, we move away from "spot-checking" Excel spreadsheets and toward robust, reproducible software engineering.

| Test Type | Question Asked | Example |
|-----------|----------------|---------|
| **Economic** | Does the math follow economic theory? | Discounting reduces value. |
| **Boundary** | Do empty inputs produce empty outputs? | Cost inputs = 0 $\rightarrow$ Total Cost = 0. |
| **Clinical** | Does the cohort flow correctly? | 100% Mortality $\rightarrow$ Min Life Expectancy. |
